{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2016-11-21 sftrends_manu.Rnw
methods draft done

%acronyms
\acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
\acrodef{din}[DIN]{dissolved inorganic nitrogen}
\acrodef{emp}[EMP]{Environmental Monitoring Program}
\acrodef{iep}[IEP]{Interagency Ecological Program}
\acrodef{jas}[JAS]{July-August-September}
\acrodef{sfe}[SFE]{San Francisco Estuary}
\acrodef{wrtds}[WRTDS]{Weighted Regressions on Time, Discharge, and Season}
\acrodef{wwtp}[WWTP]{wastewater treatment plant}

@@

Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.  These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.

@@

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \citep{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not signifcantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.  The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.  Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citep{Beck15}.

\subsection{Analysis method and application}

@@

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.  Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citealt{Beck15}). Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends. A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.  Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.  Specifically, flow-normalized trends in each analyte were summarized as both averages and percent changes from the beginning to end of annual groupings from 1976-1995, 1996-2000, 2001-2008, and 2009-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter). These annual and seasonal groupings were chosen for continuity with additional research efforts \citep[i.e.,][]{Jabusch16}.  Trends within each seasonal grouping were based on annual mean estimates for the first and last three years of flow-normalized estimates to reduce the effects of abnormal years or missing data.  Annual trends were based only on the first and last years in the time period due to limited number of years in some groupings \citepalias{BeckIP}.

@@

Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more detailed understanding of trends for further evaluation.  The stations werechosen to address ecological and management-based questions that have relevance outside of the region, having importance for our understanding of estuarine processes that influence eutrophication trends on decadal time scales.  The selected case studies focused on \begin{inparaenum}[1\upshape)]
\item nitrogen trends and flow-effects of the San Joaquin river at C10, 
\item effects of wastewater treatment upgrades upstream of P8, and
\item effects of biological invasion on nutrient dynamics in Suisun Bay.
\end{inparaenum}
Each case study is built around hypotheses that results from \ac{wrtds} models are expected to support, both as a general description and for additional testing with alternative methods.

@@

Multiple biological and physical factors influence nutrient concentrations at different temporal scales \citep[][, references therein]{Cloern10}.  As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex.  A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, is expected to reveal \begin{inparaenum}[1\upshape)]
\item an annual trend independent of a seasonal trend, and 
\item varying flow contribution to nutrient dynamics, either as a difference between predicted and flow-normalized results or changes in the nutrient versus flow relationship at different annual periods.
\end{inparaenum}
Results from this case study were used to demonstrate the breadth of information that can be obtained from the observed time series with \ac{wrtds}.

@@

Hypothesis: Modal response of nutrient concentrations at P8 over time is result of WWTP upgrades, so we expect 1) a shift in load contributions before/after upgrade, 2) a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 3) different nitrogen species will have different changes depending on change in load outputs. See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}, but this is about Tracy
\subsubsection{Effects of biological invasions}

Hypothesis: Biological invasions by benthic filter feeders have shifted abundance and composition of phytoplankton communities in Suisun Bay, so we expect 1) decline in annual, flow-normalized chlorophyll concentrations over time coincident with increase in abundance of invaders, 2) changes in ratios of limiting nutrients (nitrogen, SiO2) suggesting different uptake rates with shift in community composition, and 3) seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession.

@@

@@

Note the decrease in Potamocorbula abundance in 2011, 2012.  These are wet years where abundance/biomass of the clams is driven down by lower salinity.  Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.  The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.  This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot). See suggestions in \citet{Alpine92} regarding flow/grazer relationships in the Bay.

>>>>

%acronyms
\acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
\acrodef{din}[DIN]{dissolved inorganic nitrogen}
\acrodef{emp}[EMP]{Environmental Monitoring Program}
\acrodef{iep}[IEP]{Interagency Ecological Program}
\acrodef{jas}[JAS]{July-August-September}
\acrodef{sfe}[SFE]{San Francisco Estuary}
\acrodef{wrtds}[WRTDS]{Weighted Regressions on Time, Discharge, and Season}
\acrodef{wwtp}[WWTP]{wastewater treatment plant}
\acrodefplural{wwtp}[WWTPs]{wastewater treatment plants}

@@

Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented changes in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.  These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations \citep[e.g.,][]{Dugdale07,Parker12,Glibert14}.

@@

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \citep{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not signifcantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.  The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.  Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citep{Beck15}.  
\subsection{Analysis method and application}

@@

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.  Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citealt{Beck15}). Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends. A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.  Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.  Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter). These annual and seasonal groupings were chosen for continuity with similar comparisons reported in \citet{Jabusch16}.

Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions. This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \citep{Hirsch82}.  Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.  The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.  The results were interpreted relative to those in \citet{Jabusch16} that were based on observed data to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.  The analysis was also similar to that in \citet{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression. Functions in the EnvStats package in R were used for the seasonal Kendall tests \citep{Millard13}.

@@

Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.  The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends on decadal time scales.  The selected case studies focused on \begin{inparaenum}[1\upshape)]
\item nitrogen trends and flow-effects of the San Joaquin river at C10, 
\item effects of wastewater treatment upgrades upstream of P8, and
\item effects of biological invasion on nutrient dynamics in Suisun Bay.
\end{inparaenum}
Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.

@@

Multiple biological and physical factors influence nutrient concentrations at different temporal scales \citep[][references therein]{Cloern10}.  As a result, relationships between nutrients, time, and flow that could be described by quantitative methods like \ac{wrtds} are espected to be non-linear and complex as compared to more conventional methods for time series analysis.  A detailed evaluation of nitrogen dynamics at C10 using \ac{wrtds}, including the effects of flow, was expected to reveal \begin{inparaenum}[1\upshape)]
\item an annual trend independent of a seasonal trend, and 
\item varying flow contribution to nutrient dynamics, either as a difference between predicted and flow-normalized results or changes in the nutrient versus flow relationship at different annual periods.
\end{inparaenum}
Results from this case study were used to demonstrate the breadth of information that can be obtained from the observed time series with \ac{wrtds}.

@@

Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.  As noted in \citep{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.  Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., \citealt{Cornwell14,Novick15}).  Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}. Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in load contributions before/after upgrade,
\item a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 
\item different nitrogen species will have different changes depending on change in load outputs. 
\end{inparaenum}
The use of \ac{wrtds} to describe downstream effects of \ac{wwtp} upgrades could reveal flow-independent trends that have not been previously described.
%See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}

The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \citep{Cohen98}.  The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \citep{Carlton90,Nichols90,Werner93,Crauder16}.  In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.  Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \citep{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}.  Abundance of clams is linked to inflow rates such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.

We hypothesized that \ac{wrtds} models applied to water quality obsevations in the upper estuary would emphasize \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
\end{inparaenum}
The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.

@@

Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in \citet{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change. \cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from \citet{Jabusch16}.  Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).

@@

Note the decrease in Potamocorbula abundance in 2011, 2012.  These are wet years where abundance/biomass of the clams is driven down by lower salinity.  Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.  The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.  This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot). See suggestions in \citet{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.
